Short and Mechanized Logical Relation for Dependent Type Theory

### 2024-02-24 paper/paper.tex
Rename Coherence to convertibility

The substitution operator, which takes the form $[[a {rho}]]$,
traverses the syntax of $[[a]]$ and replaces each variable $[[var i]]$
with the term $[[rho i]]$. When traversing under binders (e.g. in the
$[[(\ a) { rho }]]$ case), the $\Uparrow$ operator prevents $[[var
0]]$, the bound variable, from being replaced by
$[[rho]]$. From the definition of $[[up rho]]$, we see that it keeps
$[[var 0]]$ unchanged during substitution and replaces each $[[var
Suc i]]$ with $[[rho i {up 1}]]$. The extra shifting with $[[up 1]]$
is required so the variables in terms from the codomain of $[[rho]]$ skip over
the newly introduced binder to refer to the correct binding
location. The substitution operator is referred to as simultaneous
substitution as it substitutes all variables at once. It is possible
to recover single substitution by composing the extension operator and
the identity substitution: $[[a { b }]] := [[a { idtm .: b }]]$.

\begin{figure}[h]
\begin{minipage}{0.9\textwidth}
\drules[P]{$[[a => b]]$}{Parallel Reduction}{AppAbs, IfTrue, IfFalse, JRefl}
\drules[PS]{$[[a =>+ b]]$}{Transitive Closure of Parallel Reduction}{Refl, Step}
\drules[C]{$[[a <=> b]]$}{Coherence}{Intro}
\end{minipage}
\caption{Parallel reduction ($\beta$-rules only) and convertibility }
\label{fig:par}
\end{figure}

% \scw{motivate parallel reduction. Why are you telling us this now. Is this
%   an explanation of definitional equality and its properties?}
Before we specify the typing rules, we need to first specify its
untyped equational theory. We choose to represent the untyped
equational theory in an algorithmic style using the notion of parallel
reduction, defined in Figure~\ref{fig:par} (congruence rules are omitted). Parallel reduction takes
the form $[[a => b]]$ and we denote its transitive as $[[a =>+
b]]$. We say that two terms $[[a0]]$ and $[[a1]]$ are convertible,
denoted by the notation $[[a0 <=> a1]]$, if there exists some term
$[[b]]$ such that $[[a0 =>+ b]]$ and $[[a1 =>+ b]]$. The symmetric
notation of convertibility suggests that it is an equivalence
relation. We sketch out the sequence of lemmas required to derive this
property below and omit the details. Our technique for proving the
properties about parallel reduction is based on
\citet{takahashi-parallel-reduction}. A modern exposition of the
same technique can be found in \citet{plfa22.08}.
\begin{lemma}[Par Refl\footnote{\dotv{join.v}{Par\_refl}}]
  \label{lemma:parrefl}
  For all terms $[[a]]$, $[[a => a]]$.
\end{lemma}
\begin{lemma}[Par cong\footnote{\dotv{join.v}{par\_cong}}]
  \label{lemma:parcong}
  If $[[a0 => a1]]$ and $[[b0 => b1]]$, then $[[a0 { b0 } => a1 { b1 }]]$.
\end{lemma}
\begin{corollary}[Par subst\footnote{\dotv{join.v}{par\_subst}}]
  \label{lemma:parsubst}
  If $[[a0 => a1]]$, then $[[a0 {b} => a1 {b}]]$ for arbitrary $[[b]]$.
\end{corollary}
\begin{lemma}[Par diamond\footnote{\dotv{join.v}{par\_confluent}}]
  \label{lemma:pardiamond}
  If $[[a => b0]]$ and $[[a => b1]]$, then there exists some term
  $[[c]]$ such that $[[b0 => c]]$ and $[[b1 => c]]$.
\end{lemma}
\begin{lemma}[Coherence refl\footnote{\dotv{join.v}{Coherent\_reflexive}}]
  \label{lemma:coherencerefl}
  For all terms $[[a]]$, $[[a <=> a]]$.
\end{lemma}
\begin{lemma}[Coherence sym\footnote{\dotv{join.v}{Coherent\_symmetric}}]
  \label{lemma:coherencesym}
  If $[[a <=> b]]$, then $[[b <=> a]]$.
\end{lemma}
\begin{lemma}[Coherence trans\footnote{\dotv{join.v}{Coherent\_transitive}}]
  \label{lemma:coherencetrans}
  If $[[a0 <=> a1]]$ and $[[a1 <=> a2]]$, then $[[a0 <=> a2]]$.
\end{lemma}
From Lemma~\ref{lemma:coherencerefl}, \ref{lemma:coherencesym}, and
\ref{lemma:coherencetrans}, we conclude that coherence is indeed an
equivalence relation. It is then trivial to show that all untyped
$\beta$-equivalence rules are admissible, though we omit this proof
from our mechanization and will from now on focus on parallel
reduction since the diamond property (Lemma~\ref{lemma:pardiamond})
makes it easier to work with. A detailed argument of the equivalence
between $[[a <=> b]]$ and untyped $\beta$-equivalence can be found in
\citet{barendregt:lambda-calculi-with-types} and
\citet{takahashi-parallel-reduction}.

\begin{figure}[h]
\begin{minipage}{0.9\textwidth}
\drules[Ctx]{$[[ |- G]]$}{Context Well-Formedness}{Empty, Cons}
\drules[T]{$[[G |-  a : A]]$}{Typing}{Var, Set, Pi, Abs, App, Conv,
  Refl, Eq, J, If, Bool, True, False, Void}
\end{minipage}
\caption{Syntactic typing for \lang}
\label{fig:typing}
\end{figure}
% Figure~\ref{fig:par} shows the definition of the parallel reduction
% relation, which takes the form $[[a => b]]$. We use $[[a =>+ b]]$ to
% represent its transitive and reflexive closure, which in turn allows us to define
% the coherence relation $[[a <=> b]]$.\scw{where is the terminology ``coherence''
% from? I'm not familiar with it. Maybe you can replace it with ``definitional
% equivalence'' as that is how it is used in \lang?}
% We say that two terms $[[a]]$
% and $[[b]]$ are coherent if they can eventually reduce to some common
% term $[[c]]$ through parallel reduction. The symmetric notation of
% coherence suggests\scw{but you prove it below?} it is an equivalence relation.

% We sketch out some key properties about parallel reduction and
% coherence without giving their proofs. Our technique for establishing
% those results is based on

% \scw{cite barendregt?}
% \yl{resolved}

% TODO: remove the lengthy discussion below

% Now, we prove that coherence is indeed an equivalence relation.

% First, we show that coherence is reflexive through the following
% sequence of lemmas.

% Lemma~\ref{lemma:parrefl} can be proven by structural induction over
% the term $[[a]]$. Lemmas~\ref{lemma:parsrefl} and \ref{lemma:coherencerefl}
% immediately follow as corollaries of Lemma~\ref{lemma:parrefl}.

% The reflexivity of parallel reduction enables us to embed rules from
% call-by-name semantics into parallel reduction, as the following lemma
% shows.
% \begin{lemma}[Par AbsCbn\footnote{\dotv{join.v}{P\_AppAbs\_cbn}}] For all $[[A]], [[a]],$ and $[[b]]$,
%   \label{lemma:parabscbn}
%   $[[(\ A a) b => b {a}]]$
% \end{lemma}
% \begin{proof}
%   Immediate from Lemma~\ref{lemma:parrefl} and \rref{P-AppAbs}.
% \end{proof}

% Symmetry of coherence immediately falls from its definition.

% Before we can prove transitivity, we need to show that parallel
% reduction satisfies the diamond property.

% The congruence property (Lemma~\ref{lemma:parcong}) can be proven by
% structural induction over the derivation of $[[a0 => a1]]$.
% Likewise, Lemma~\ref{lemma:pardiamond} can be proven by structural induction
% over the derivation of $[[a => b0]]$. The \rref{P-AppAbs} case requires
% the use of Lemma~\ref{lemma:parcong}.

% From Lemma~\ref{lemma:parcong} and \ref{lemma:parrefl}, we recover the
% single substitution property as a simple corollary.

% A relation that satisfies the
% diamond property must also be confluent, meaning that its transitive
% and reflexive closure is confluent.
% \begin{lemma}[Par confluent\footnote{\dotv{join.v}{pars\_confluent}}]
%   \label{lemma:parconfluent}
%   If $[[a =>+ b0]]$ and $[[a =>+ b1]]$, then there exists some term
%   $[[c]]$ such that $[[b0 =>+ c]]$ and $[[b1 =>+ c]]$.
% \end{lemma}
% While $[[a =>+ b]]$ is defined as the transitive closure of $[[a => b]]$,
% it coincides with the transitive and reflexive closure of $[[a => b]]$ since $[[a => b]]$ is reflexive (Lemma~\ref{lemma:parrefl}).

% The transitivity of the coherence relation follows as a corollary of
% Lemma~\ref{lemma:parconfluent}.
% By the definition of coherence, there exists some term $[[b]]$ such that $[[a0 =>+ b0]]$,
% $[[a1 =>+ b0]]$ and some term $[[b1]]$ such that $[[a1 =>+ b1]]$ and
% $[[a2 =>+ b1]]$. By Lemma~\ref{lemma:parconfluent}, there exists some
% term $[[c]]$ such that $[[b0 =>+ c]]$ and $[[b1 =>+ c]]$. It sufficies
% to show that $[[a0 =>+ c]]$ and $[[a2 =>+ c]]$, both of which
% trivially hold since the transitive closure $[[a =>+ b]]$ is transitive.
% This concludes the proof that coherence is an equivalence relation.
% \begin{lemma}[Coherence Equivalence]
%   \label{lemma:coherenceequiv}
%   The relation $[[a <=> b]]$ satisfies reflexivity, symmetry, and
%   transitivity and therefore is an equivalence relation.
% \end{lemma}

Figure~\ref{fig:typing} gives the full typing rules for
\lang{}. The premises wrapped in \colorbox{lightgray}{gray} boxes can be shown to be
admissible syntactically, though some of them are required to
strengthen the inductive hypothesis of the fundamental theorem.
% \scw{Is the precondition for T-ABS really admissible? Isn't it required
% to ensure predicativity? Have you proved it? Have you proved regularity?}
% \yl{I proved regularity. \Rref{T-Abs} is admissible if cumulativity
%   was formulated correctly (which wasn't the case in the version
%   of the paper you read). I revised \rref{T-Univ} so the premise is
%   not admissible }
% \scw{Do you have a formation rule for identity types?}
% \yl{Added}

@@

\citet{siles2012pure} shows the equivalence of Barendregt's
Pure Type System~\citep{barendregt1991introduction}, which employs
untyped equality, and its variant that uses typed judgmental
equality. This assures us that we do not lose much generality working
with a system with untyped conversion. We include a detailed
discussion in Section~\ref{sec:relatedwork} on how type-directed and
untyped equality affect the applicability of our technique.

@@

% It is trivial to embed a system with judgmental
% equality to a system with untyped equality by erasing typing
% information. As a result, it is easy to port our consistency result from our
% type system to a variant with typed judgmental equality
% as long as the typed system does not include $\eta$ laws
% that would require type annotations (e.g. the $\eta$-law for unit
% types).
% and we discuss in Section~\ref{sec:extension} how we can extend
% our proof to handle those rules.
\yl{Unresolved: is it ok to say our system's conversion rule is untyped like
Coq?}
% \scw{Maybe move more of this paragraph and the next to the discussion/related
%   work section, and add a forward pointer here.}
% \scw{Just say that like Coq, the definition of the system uses untyped
% equality}

@@

\section{Logical Relation}
% \scw{This section pre-supposes that we want to define a logical relation,
% but doesn't precisely state what we want to use it to prove, and why a logical
% relation is suitable. (And why the property you want to prove is difficult to
% show!) Should add more motivation here.}
\scw{We need to explicitly point out that the key ideas of this paper are
  discussed, here, in this section.
  We need to explicitly remark on why logical relations are difficult to
  define for dependent type theory and explain why this setting is more
  difficult than with simple types (STLC) or with polymorphic types (System F).
  \begin{itemize}
  \item Dependent function types (result type mentions argument value, so can't define
    relation by recursion over type structure)
  \item Definitional equality (not all types look like types)
  \item Identity types(?)
  \end{itemize}
  Should we be more explicit in our comparison with Girard's trick for polymorphic type?
  There, the definition stays recursive because it doesn't substitute for the variables
  in the function types. But that approach is not available in this setting, because not
  all quantified things are types. And we might need that information to interpret, say,
  identity types in the right way.
}
\scw{ We also need to explicitly point out that our logical relation is untyped.
  This has two benefits: it allows semantic typing to be meaningful independent from
  syntactic typing (cite Derek, forward reference to next section) and it avoids
  significant bookkeeping, especially in the case of Kripke logical relations (we need to define
  what these are).
  Is there a cost to an untyped relation?
}
Before we define our logical relation, we first formally specify the
consistency property that we want to prove.
\begin{theorem}[Logical Consistency]
  \label{theorem:consistency}
  The judgment $[[empty |- a : Void ]]$ is not derivable.
\end{theorem}
The property can be formulated in a simply typed language, where
$[[Void]]$ is similarly defined as a type that has no term. A related
property, referred to as the termination property (for closed terms),
is commonly used in introductory materials such as
\citet{skorstengaard2019introduction}, \citet{pierce2002types}, and
\citet{harpertait} to motivate the need for a logical relation.

@@

\section{Existence of $\beta\eta$-normal forms}
We can easily incorporate the function $\eta$ law to the equational
theory of \lang{} by adding the following parallel reduction rule.
\begin{center}
  \drule[width=2.5in]{P-AbsEta}
\end{center}
% \Rref{P-AbsEta} effectively adds the $\eta$ law for functions to our
% equational theory since convertibility ($[[a <=> b]]$), the untyped relation
% used for type conversion, is built on top of parallel reduction.
We can recover the same confluence result about parallel reduction
using the standard techniques from
\citet{barendregt:lambda-calculi-with-types,
takahashi-parallel-reduction}, though anti-renaming
(Lemma~\ref{lemma:parantirenaming}) needs to be proven before the
diamond property (Lemma~\ref{lemma:pardiamond}). Another complication
is that confluence and anti-renaming now requires induction on a size
metric of lambda terms so the $\eta$ case has a useful induction hypothesis.

>>>>

The substitution operator, which takes the form $[[a {rho}]]$,
traverses the syntax of $[[a]]$ and replaces each variable $[[var i]]$
with the term $[[rho i]]$. When traversing under binders (e.g. in the
$[[(\ a) { rho }]]$ case), the $\Uparrow$ operator prevents $[[var
0]]$, the bound variable, from being replaced by
$[[rho]]$. From the definition of $[[up rho]]$, we see that it keeps
$[[var 0]]$ unchanged during substitution and replaces each $[[var
Suc i]]$ with $[[rho i {up 1}]]$. The extra shifting with $[[up 1]]$
is required so the variables in terms from the codomain of $[[rho]]$ skip over
the newly introduced binder to refer to the correct binding
location. The substitution operator is referred to as simultaneous
substitution as it substitutes all variables at once. It is possible
to recover single substitution by composing the extension operator and
the identity substitution: $[[a { b }]] := [[a { idtm .: b }]]$. Since
the definition of semantic typing in Section~\ref{sec:logrelproof}
relies on simultaneous substitution, we find it more convenient to
formulate simultaneous substitution directly rather than recovering it
from single substitution.

\begin{figure}[h]
\begin{minipage}{0.9\textwidth}
\drules[P]{$[[a => b]]$}{Parallel Reduction}{AppAbs, IfTrue, IfFalse, JRefl}
\drules[PS]{$[[a =>+ b]]$}{Transitive Closure of Parallel Reduction}{Refl, Step}
\drules[C]{$[[a <=> b]]$}{Convertibility}{Intro}
\end{minipage}
\caption{Parallel reduction ($\beta$-rules only) and convertibility }
\label{fig:par}
\end{figure}

% \scw{motivate parallel reduction. Why are you telling us this now. Is this
%   an explanation of definitional equality and its properties?}
Before we specify the typing rules, we need to first specify its
untyped equational theory. We choose to represent the untyped
equational theory in an algorithmic style using the notion of parallel
reduction, defined in Figure~\ref{fig:par} (congruence rules are omitted).
 Parallel reduction takes
the form $[[a => b]]$ and we denote its transitive as $[[a =>+
b]]$. We say that two terms $[[a0]]$ and $[[a1]]$ are convertible,
denoted by the notation $[[a0 <=> a1]]$, if there exists some term
$[[b]]$ such that $[[a0 =>+ b]]$ and $[[a1 =>+ b]]$. The symmetric
notation of convertibility suggests that it is an equivalence
relation. We sketch out the sequence of lemmas required to derive this
property below and omit the details. Our technique for proving the
properties about parallel reduction is based on
\citet{takahashi-parallel-reduction}. A modern exposition of the
same technique can be found in \citet{plfa22.08}.
\begin{lemma}[Par Refl\footnote{\dotv{join.v}{Par\_refl}}]
  \label{lemma:parrefl}
  For all terms $[[a]]$, $[[a => a]]$.
\end{lemma}
\begin{lemma}[Par cong\footnote{\dotv{join.v}{par\_cong}}]
  \label{lemma:parcong}
  If $[[a0 => a1]]$ and $[[b0 => b1]]$, then $[[a0 { b0 } => a1 { b1 }]]$.
\end{lemma}
\begin{corollary}[Par subst\footnote{\dotv{join.v}{par\_subst}}]
  \label{lemma:parsubst}
  If $[[a0 => a1]]$, then $[[a0 {b} => a1 {b}]]$ for arbitrary $[[b]]$.
\end{corollary}
\begin{lemma}[Par diamond\footnote{\dotv{join.v}{par\_confluent}}]
  \label{lemma:pardiamond}
  If $[[a => b0]]$ and $[[a => b1]]$, then there exists some term
  $[[c]]$ such that $[[b0 => c]]$ and $[[b1 => c]]$.
\end{lemma}
\begin{lemma}[Convertibility refl\footnote{\dotv{join.v}{Coherent\_reflexive}}]
  \label{lemma:coherencerefl}
  For all terms $[[a]]$, $[[a <=> a]]$.
\end{lemma}
\begin{lemma}[Convertibility sym\footnote{\dotv{join.v}{Coherent\_symmetric}}]
  \label{lemma:coherencesym}
  If $[[a <=> b]]$, then $[[b <=> a]]$.
\end{lemma}
\begin{lemma}[Convertibility trans\footnote{\dotv{join.v}{Coherent\_transitive}}]
  \label{lemma:coherencetrans}
  If $[[a0 <=> a1]]$ and $[[a1 <=> a2]]$, then $[[a0 <=> a2]]$.
\end{lemma}
From Lemma~\ref{lemma:coherencerefl}, \ref{lemma:coherencesym}, and
\ref{lemma:coherencetrans}, we conclude that convertibility is indeed an
equivalence relation. It is then trivial to show that all untyped
$\beta$-equivalence rules are admissible, though we omit this proof
from our mechanization and will from now on focus on parallel
reduction since the diamond property (Lemma~\ref{lemma:pardiamond})
makes it easier to work with. A detailed argument of the equivalence
between $[[a <=> b]]$ and untyped $\beta$-equivalence can be found in
\citet{barendregt:lambda-calculi-with-types} and
\citet{takahashi-parallel-reduction}.

Untyped equality allows us to show the
equivalence between our more structured notion of convertibility
 ($[[a <=> b]]$) and the declarative, untyped $\beta$-equivalence
independently from the type system and the logical relation through
well-established syntactic approaches. The directedness of the
convertibility relation will be leveraged in
Section~\ref{sec:logreldep} to justify the conversion rule of our
typing judgment, which we will introduce shortly.

\begin{figure}[h]
\begin{minipage}{0.9\textwidth}
\drules[Ctx]{$[[ |- G]]$}{Context Well-Formedness}{Empty, Cons}
\drules[T]{$[[G |-  a : A]]$}{Typing}{Var, Set, Pi, Abs, App, Conv,
  Refl, Eq, J, If, Bool, True, False, Void}
\end{minipage}
\caption{Syntactic typing for \lang}
\label{fig:typing}
\end{figure}

Figure~\ref{fig:typing} gives the full typing rules for
\lang{}. The premises wrapped in \colorbox{lightgray}{gray} boxes can be shown to be
admissible syntactically, though some of them are required to
strengthen the inductive hypothesis of the fundamental theorem.

@@

\citet{siles2012pure} shows the equivalence of Barendregt's
Pure Type System~\citep{barendregt1991introduction}, which employs
untyped equality, and its variant that uses typed judgmental
equality. This assures us that we do not lose much generality working
with a system with untyped conversion.
We include a detailed discussion in Section~\ref{sec:discuss} on how
type-directed and untyped equality affect the applicability of our
technique.

@@

% It is trivial to embed a system with judgmental
% equality to a system with untyped equality by erasing typing
% information. As a result, it is easy to port our consistency result from our
% type system to a variant with typed judgmental equality
% as long as the typed system does not include $\eta$ laws
% that would require type annotations (e.g. the $\eta$-law for unit
% types).
% and we discuss in Section~\ref{sec:extension} how we can extend
% our proof to handle those rules.
% \scw{Maybe move more of this paragraph and the next to the discussion/related
%   work section, and add a forward pointer here.}
% \scw{Just say that like Coq, the definition of the system uses untyped
% equality}

@@

\section{Logical Relation}
% \scw{This section pre-supposes that we want to define a logical relation,
% but doesn't precisely state what we want to use it to prove, and why a logical
% relation is suitable. (And why the property you want to prove is difficult to
% show!) Should add more motivation here.}
\scw{We need to explicitly point out that the key ideas of this paper are
  discussed, here, in this section.
  We need to explicitly remark on why logical relations are difficult to
  define for dependent type theory and explain why this setting is more
  difficult than with simple types (STLC) or with polymorphic types (System F).
  \begin{itemize}
  \item Dependent function types (result type mentions argument value, so can't define
    relation by recursion over type structure)
  \item Definitional equality (not all types look like types)
  \item Identity types(?)
  \end{itemize}
  Should we be more explicit in our comparison with Girard's trick for polymorphic type?
  There, the definition stays recursive because it doesn't substitute for the variables
  in the function types. But that approach is not available in this setting, because not
  all quantified things are types. And we might need that information to interpret, say,
  identity types in the right way.
}
\yl{I don't think identity types and dependent function types should
  be listed as reasons why the logrel can't be a recursive
  function. System DE has both dependent function types and identity
  types but the logical fragment doesn't have type-level computation
  so it can be defined structurally by specifying some kind of size metric
}
\scw{ We also need to explicitly point out that our logical relation is untyped.
  This has two benefits: it allows semantic typing to be meaningful independent from
  syntactic typing (cite Derek, forward reference to next section) and it avoids
  significant bookkeeping, especially in the case of Kripke logical relations (we need to define
  what these are).
  Is there a cost to an untyped relation?
}

Before we define our logical relation, we first formally specify the
consistency property that we want to prove.
\begin{theorem}[Logical Consistency]
  \label{theorem:consistency}
  The judgment $[[empty |- a : Void ]]$ is not derivable.
\end{theorem}
The property can be formulated in a simply typed language, where
$[[Void]]$ is similarly defined as a type that has no term. A related
property, referred to as the termination property (for closed terms),
is commonly used in introductory materials such as
\citet{skorstengaard2019introduction}, \citet{pierce2002types}, and
\citet{harpertait} to motivate the need for a logical relation.

@@

\section{Existence of $\beta\eta$-normal forms}
\label{sec:betaeta}
We can easily incorporate the function $\eta$ law to the equational
theory of \lang{} by adding the following parallel reduction rule.
\begin{center}
  \drule[width=2.5in]{P-AbsEta}
\end{center}
% \Rref{P-AbsEta} effectively adds the $\eta$ law for functions to our
% equational theory since convertibility ($[[a <=> b]]$), the untyped relation
% used for type conversion, is built on top of parallel reduction.
We can recover the same confluence result about parallel reduction
using the standard techniques from
\citet{barendregt:lambda-calculi-with-types,
takahashi-parallel-reduction}, though anti-renaming
(Lemma~\ref{lemma:parantirenaming}) needs to be proven before the
diamond property (Lemma~\ref{lemma:pardiamond}). Another complication
is that confluence and anti-renaming now requires induction on a size
metric of lambda terms so the $\eta$ case has a useful induction hypothesis.
